A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-16 03-semantics.tex
Reformatação do final da Seção 3.2
Reformatei as regras e definições de funções restantes.  Mantive os
originais em comentários.

The~$\nst$ transition rules for atomic expressions are defined as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL mem(id), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(mem)}        \\
%%%
\LL emit(id), n, \epsilon \RR &\NST
\LL @canrun(n), n, id \RR
    & \textbf{(emitInt)}    \\
%%%
\LL @canrun(n), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(canrun)}     \\
\end{align*}
}
%
A $mem$ operation becomes a $@nop$ which indicates the memory access (rule
\textbf{mem}).
An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
can only resume at level $n$ (rule \textbf{emitInt}).
Since all $\NST$ rules can only transit with $e=\epsilon$, an $emitInt$ causes
rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
the stack.
Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
transit, providing the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
          \LL p, n, \epsilon \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
          \LL q,n,\epsilon \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~;~q), n, \epsilon \RR \NST \LL (p'~;~q), n, e \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL (@nop~;~q),n,\epsilon \RR \NST  \LL q,n,\epsilon \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL (break~;~q),n,\epsilon \RR \NST \LL break,n,\epsilon \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
Function $val$ receives a memory identifier and returns its current value.
%
%Although the value here is arbitrary, it is unique in a reaction, because a 
%given expression can execute only once within it (remember that $loops$ must 
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

The rules for loops are analogous to sequences, but use \code{`@'} as 
separators to properly bind breaks to their enclosing loops:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (loop~p),n,\epsilon \RR \NST \LL (p~@loop~p), n, \epsilon \RR
    & \textbf{(loop-expd)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
% -----------------------------------------------------------
    { \DS \LL (p~@loop~q),n,\epsilon \RR \NST \LL (p'~@loop~q), n, e \RR }
    & \textbf{(loop-adv)}    \\
%%%
& \LL (@nop~@loop~p), n, \epsilon \RR \NST \LL (loop~p), n, \epsilon \RR
    & \textbf{(loop-nop)}    \\
%%%
& \LL (break~@loop~p), n, \epsilon \RR \NST \LL @nop, n, \epsilon \RR
    & \textbf{(loop-brk)}
\end{eqnarray*}
}

%
When a program encounters a $loop$, it first expands its body in sequence with 
itself (rule \textbf{loop-expd}).
Rules \textbf{loop-adv} and \textbf{loop-nop} are similar to rules 
\textbf{seq-adv} and \textbf{seq-nop}, advancing the loop until they reach a 
$@nop$.
However, what follows the loop is the loop itself (rule \textbf{loop-nop}).
Note that if we used \code{`;'} as a separator in loops, rules 
\textbf{loop-brk} and \textbf{seq-brk} would conflict.
%
Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything into 
a $@nop$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (p~and~q),n,\epsilon \RR \NST \LL (p~@and~(@canrun(n)~;~q)),n,\epsilon \RR
    & \textbf{(and-expd)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~q),n,\epsilon \NST \LL (p'~@and~q),n,e \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~q),n,\epsilon \RR \NST \LL (p~@and~q'), n, e \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \LL (p~or~q), n, \epsilon \RR \NST \LL (p~@or~(@canrun(n)~;~q)), n, \epsilon \RR
    & \textbf{(or-expd)}       \\
%%%
& \frac
    { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p'~@or~q), n, e \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p~@or~q'), n, e \RR }
    & \textbf{(or-adv2)}   %\\
\end{eqnarray*}
}
%
Rules \textbf{and-expd} and \textbf{or-expd} insert a $@canrun(n)$ at the
beginning of the right branch.
This ensures that an $emitInt$ on the left branch, which transits to a
$@canrun(n)$, still resumes before the right branch starts.
%
The deterministic behavior of the semantics relies on the \emph{isblocked} 
predicate (see Figure~\ref{fig.isblocked}) which appears in rules
\textbf{and-adv2} and \textbf{or-adv2}.
These rules require the left branch $p$ to be blocked in order to allow the 
right branch transition from $q$ to $q'$.

For a parallel $@and$, if one of the sides terminates, the composition is
simply substituted by the other side (rules \textbf{and-nop1} and
\textbf{and-nop2}.
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
    & \textbf{(and-nop1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR }
    & \textbf{(and-nop2)}   \\
%%%
& \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}
%
The $clear$ function (see Figure~\ref{fig.formal.clear}) concatenates all
active $fin$ bodies of the side being aborted, so that they execute before the
composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
%the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that $fin$ bodies cannot
contain awaiting expressions ($awaitExt$, $awaitInt$, $every$, and $fin$),
i.e., they are guaranteed to execute entirely within a reaction.

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    { \DS isblocked(n,p) }
%   -----------------------------------------------------------
    { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaitExt$, $awaitInt$, $every$, $fin$, and
$@canrun$ expressions in parallel trails.
%
If no trails are blocked in $@canrun$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $@canrun$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \textbf{pop}).

\begin{figure}
{\small
\begin{align*}
  bcast(e, awaitExt(e)) &= @nop                         \\
  bcast(e, awaitInt(e)) &= @nop                         \\
  bcast(e, every~e~p)   &= p;~every~e~p                 \\
  bcast(e, @canrun(n))  &= @canrun(n)                   \\
  bcast(e, fin~p)       &= fin~p                        \\
  bcast(e, p~;~q)       &= bcast(e,p)~;~q               \\
  bcast(e, p~@loop~q)   &= bcast(e,p)~@loop~q           \\
  bcast(e, p~@and~q)    &= bcast(e,p)~@and~bcast(e,q)   \\
  bcast(e, p~@or~q)     &= bcast(e,p)~@or~bcast(e,q)    \\
  bcast(e, \_)          &= \bot \2 (mem,emitInt,break,if,  \\
                                 & \5\5 loop,and,or,@nop) %\\
\end{align*}
}
\caption{
The function $bcast$ awakes awaiting trails matching the event by converting
$awaitExt$ and $awaitInt$ to $@nop$ expressions, and by unwinding $every$
expressions.
\label{fig.bcast}
}
\end{figure}

\begin{figure}
{\small
\begin{align*}
  isblocked(n, \1 awaitExt(id)) &= true                                   \\
  isblocked(n, \1 awaitInt(id)) &= true                                   \\
  isblocked(n, \1 every~e~p)    &= true                                   \\
  isblocked(n, \1 @canrun(m))   &= (n > m)                                \\
  isblocked(n, \1 fin~p)        &= true                                   \\
  isblocked(n, \1 p~;~q)        &= isblocked(n,p)                         \\
  isblocked(n, \1 p~@loop~q)    &= isblocked(n,p)                         \\
  isblocked(n, \1 p~@and~q)     &= isblocked(n,p) \wedge isblocked(n,q)   \\
  isblocked(n, \1 p~@or~q)      &= isblocked(n,p) \wedge isblocked(n,q)   \\
  isblocked(n, \1 \_)           &= false \2 (mem,emitInt,break,if,        \\
                                & \5\5\5\1 loop,and,or,@nop)   %\\
\end{align*}
}
\caption{
The predicate $isblocked$ is true only if all branches in parallel are blocked
waiting for events, finalization clauses, or certain stack levels.
\label{fig.isblocked}
}
\end{figure}

\begin{figure}[b]
{\small
\begin{align*}
  clear( awaitExt(e) ) &= @nop                  \\
  clear( awaitInt(e) ) &= @nop                  \\
  clear( every~e~p )   &= @nop                  \\
  clear( @canrun(n) )  &= @nop                  \\
  clear( fin~p )       &= p                     \\
  clear( p~;~q )       &= clear(p)              \\
  clear( p~@loop~q )   &= clear(p)              \\
  clear( p~@and~q )    &= clear(p)~;~clear(q)   \\
  clear( p~@or~q )     &= clear(p)~;~clear(q)   \\
  clear( \_ )          &= \bot \2 (mem,emitInt,break,if, \\
                                  & \5\5 loop,and,or,@nop) %\\ 
\end{align*}
}
\caption{
The function $clear$ extracts $fin$ expressions in parallel and put their 
bodies in sequence.
\label{fig.formal.clear}
}
\end{figure}

>>>>

The~$\nst$ transition rules for atomic expressions are defined as follows:
\begin{align*}
  \<\ceu{\Mem(\Id)},n,\nil>
  &\nst\<\ceu{\Nop},n,\nil>\Rtag{mem}\\
  %%
  \<\ceu{\EmitInt(\Id)},n,\nil>
  &\nst\<\ceu{\CanRun(n)},n,\ceu{\Id}>\Rtag{emit-int}\\
  %%
  \<\ceu{\CanRun(n)},n,\nil>
  &\nst\<\ceu{\Nop},n,\nil>\Rtag{can-run}
\end{align*}

%-
% { \setlength{\jot}{20pt}
% \begin{align*}
% \LL mem(id), n, \epsilon \RR &\NST
% \LL @nop, n, \epsilon \RR
%     & \textbf{(mem)}        \\
% %%%
% \LL emit(id), n, \epsilon \RR &\NST
% \LL @canrun(n), n, id \RR
%     & \textbf{(emitInt)}    \\
% %%%
% \LL @canrun(n), n, \epsilon \RR &\NST
% \LL @nop, n, \epsilon \RR
%     & \textbf{(canrun)}     \\
% \end{align*}
% }
%-

A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory access (rule
\R{mem}).
An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a $\ceu{\CanRun(n)}$ which
can only resume at level~$n$ (rule \R{emit-int}).
Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$ causes
rule \R{push} to execute at the outer level, creating a new level~$n+1$ on
the stack.
Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
transit (rule~\R{can-run}), providing the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
\begin{gather*}
  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{if-true}\\[2\jot]
  %%
  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
  \DisplayProof
  \Rtag{if-false}\\[2\jot]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
  \DisplayProof
  \Rtag{seq-adv}
\end{gather*}
\vskip-\belowdisplayskip
\vskip-\abovedisplayskip
\vskip2\jot
\begin{align*}
  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[2\jot]
  %%
  \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
\end{align*}

%-
% { \setlength{\jot}{20pt}
% \begin{eqnarray*}
% & \frac
%     { \DS val(id) \neq 0 }
% %   -----------------------------------------------------------
%     { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
%           \LL p, n, \epsilon \RR }
%     & \textbf{(if-true)}       \\
% %%%
% & \frac
%     { \DS val(id,n) = 0 }
% %   -----------------------------------------------------------
%     { \DS \LL (if~mem(id)~then~p~else~q),n,\epsilon \RR \NST
%           \LL q,n,\epsilon \RR }
%     & \textbf{(if-false)}       \\
% %%%
% & \frac
%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
% %   -----------------------------------------------------------
%     { \DS \LL (p~;~q), n, \epsilon \RR \NST \LL (p'~;~q), n, e \RR }
%     & \textbf{(seq-adv)}      \\
% %%%
% & \LL (@nop~;~q),n,\epsilon \RR \NST  \LL q,n,\epsilon \RR
%     & \textbf{(seq-nop)}      \\
% %%%
% & \LL (break~;~q),n,\epsilon \RR \NST \LL break,n,\epsilon \RR
%     & \textbf{(seq-brk)}
% \end{eqnarray*}
% }
%-

Given that our semantics focuses on control, rules \R{if-true} and
\R{if-false} are the only to query~$\ceu{\Mem}$ expressions.
%
Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression.
%
%Although the value here is arbitrary, it is unique in a reaction, because a 
%given expression can execute only once within it (remember that $loops$ must 
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

The rules for loops are analogous to sequences, but use ``\code{@}'' as
separators to properly bind breaks to their enclosing loops:
\begin{align*}
  \<\ceu{\Loop{p}},n,\nil>
  &\nst\<\ceu{p\AtLoop{p}},n,\nil>\Rtag{loop-expd}\\[2\jot]
  %%
  &\hskip-6.35em
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtLoop{p}},n,\nil>\nst\<\ceu{p'\AtLoop{p}},n,e>$}
  \DisplayProof
  \Rtag{loop-adv}\\[2\jot]
  %%
  \<\ceu{\Nop\AtLoop{p}},n,\nil>
  &\nst\<\ceu{\Loop{p}},n,\nil>\Rtag{loop-nop}\\[2\jot]
  %%
  \<\ceu{\Break\AtLoop{p}},n,\nil>
  &\nst\<\ceu{\Nop},n,\nil>\Rtag{loop-brk}
\end{align*}

%-
% %
% { \setlength{\jot}{20pt}
% \begin{eqnarray*}
% & \LL (loop~p),n,\epsilon \RR \NST \LL (p~@loop~p), n, \epsilon \RR
%     & \textbf{(loop-expd)}       \\
% %%%
% & \frac
%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
% % -----------------------------------------------------------
%     { \DS \LL (p~@loop~q),n,\epsilon \RR \NST \LL (p'~@loop~q), n, e \RR }
%     & \textbf{(loop-adv)}    \\
% %%%
% & \LL (@nop~@loop~p), n, \epsilon \RR \NST \LL (loop~p), n, \epsilon \RR
%     & \textbf{(loop-nop)}    \\
% %%%
% & \LL (break~@loop~p), n, \epsilon \RR \NST \LL @nop, n, \epsilon \RR
%     & \textbf{(loop-brk)}
% \end{eqnarray*}
% }
%-

When a program encounters a $\ceu{\Loop}$, it first expands its body in sequence with 
itself (rule \R{loop-expd}).
Rules \R{loop-adv} and \R{loop-nop} are similar to rules 
\R{seq-adv} and \R{seq-nop}, advancing the loop until a~$\ceu{\Nop}$ is reached.
However, what follows the loop is the loop itself (rule \R{loop-nop}).
Note that if we used ``\code{;}'' as a separator in loops, rules
\R{loop-brk} and \R{seq-brk} would conflict.
%
Rule \R{loop-brk} escapes the enclosing loop, transforming everything into 
a~$\ceu{\Nop}$.
%Rule \textbf{loop-brk} escapes the enclosing loop, transforming everything 
%into a $clear(p)$.
%We cannot simply transform the loop into a $nop$ because its body may be a 
%parallel composition containing finalization blocks.

The semantic rules for~$\ceu{\And}$ and~$\ceu{\Or}$ compositions
force transitions on their left branches~$p$ to occur before their right
branches~$q$:
\begin{gather*}
  \hskip-.5em
  \<\ceu{p\And{q}},n,\nil>
  \nst\<\ceu{p\And(\CanRun(n);q)},n,\nil>
  \Rtag{and-expd}\\[2\jot]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p'\AtAnd{q}},n,e>$}
  \DisplayProof
  \Rtag{and-adv1}\\[2\jot]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtAnd{q}},n,\nil>\nst\<\ceu{p\AtAnd{q'}},n,e>$}
  \DisplayProof
  \Rtag{and-adv2}\\[2\jot]
  %%
  \<\ceu{p\Or{q}},n,\nil>
  \nst\<\ceu{p\Or(\CanRun(n);q)},n,\nil>
  \Rtag{or-expd}\\[2\jot]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p'\AtOr{q}},n,e>$}
  \DisplayProof
  \Rtag{or-adv1}\\[2\jot]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \AxiomC{$\<q,n,\nil>\nst\<q',n,e>$}
  \BinaryInfC{$\<\ceu{p\AtOr{q}},n,\nil>\nst\<\ceu{p\AtOr{q'}},n,e>$}
  \DisplayProof
  \Rtag{or-adv2}
\end{gather*}

%-
% { \setlength{\jot}{20pt}
% \begin{eqnarray*}
% & \LL (p~and~q),n,\epsilon \RR \NST \LL (p~@and~(@canrun(n)~;~q)),n,\epsilon \RR
%     & \textbf{(and-expd)}       \\
% %%%
% & \frac
%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@and~q),n,\epsilon \NST \LL (p'~@and~q),n,e \RR }
%     & \textbf{(and-adv1)}      \\
% %%%
% & \frac
%     { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@and~q),n,\epsilon \RR \NST \LL (p~@and~q'), n, e \RR }
%     & \textbf{(and-adv2)}      \\
% %%%
% & \LL (p~or~q), n, \epsilon \RR \NST \LL (p~@or~(@canrun(n)~;~q)), n, \epsilon \RR
%     & \textbf{(or-expd)}       \\
% %%%
% & \frac
%     { \DS \LL p,n,\epsilon \RR \NST \LL p',n,e \RR }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p'~@or~q), n, e \RR }
%     & \textbf{(or-adv1)}   \\
% %%%
% & \frac
%     { \DS isblocked(n,p) \1,\2 \LL q,n,\epsilon \RR \NST \LL q',n,e \RR }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@or~q),n,\epsilon \RR \NST \LL (p~@or~q'), n, e \RR }
%     & \textbf{(or-adv2)}   %\\
% \end{eqnarray*}
% }
%-

Rules \R{and-expd} and \R{or-expd} insert a~$\ceu{\CanRun(n)}$ at the
beginning of the right branch.
This ensures that~any $\ceu{\EmitInt}$ on the left branch, which transits to a~$\ceu{\CanRun(n)}$, still resumes before the right branch starts.
%
The deterministic behavior of the semantics relies on the \emph{isblocked} 
predicate (see Figure~\ref{fig.isblocked}) which appears in rules
\R{and-adv2} and \R{or-adv2}.
These rules require the left branch~$p$ to be blocked for the
right branch to transition from~$q$ to~$q'$.

In a parallel~$\ceu{\AtAnd}$, if one of the sides terminates, the composition is
simply substituted by the other side (rules \R{and-nop1} and
\R{and-nop2} below).
%
In a parallel~$\ceu{\AtOr}$, however, if one of the sides terminates, the whole composition
terminates and function~$\clear$ is used to properly finalize the aborted
side (rules \R{or-nop1} and \R{or-nop2}).
\begin{gather*}
  \<\ceu{{\Nop}\AtAnd{q}},n,\nil>\nst\<q,n,\nil>\Rtag{and-nop1}\\[2\jot]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtAnd{\Nop}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{and-nop2}\\[2\jot]
  %%
  \<\ceu{{\Nop}\AtOr{q}},n,\nil>\nst\<\clear(q),n,\nil>\Rtag{or-nop1}\\[2\jot]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtOr{\Nop}},n,\nil>\nst\<\clear(p),n,\nil>$}
  \DisplayProof
  \Rtag{or-nop2}
\end{gather*}

%-
% { \setlength{\jot}{20pt}
% \begin{eqnarray*}
% & \LL (@nop~@and~q), n, \epsilon \RR \NST \LL q,n,\epsilon \RR
%     & \textbf{(and-nop1)}   \\
% %%%
% & \frac
%     { \DS isblocked(n,p) }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@and~@nop), n, \epsilon \RR \NST \LL p,n,\epsilon \RR }
%     & \textbf{(and-nop2)}   \\
% %%%
% & \LL (@nop~@or~q), n, \epsilon \RR \NST \LL clear(q),n,\epsilon \RR
%     & \textbf{(or-nop1)}   \\
% %%%
% & \frac
%     { \DS isblocked(n,p) }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@or~@nop), n, \epsilon \RR \NST \LL clear(p),n,\epsilon \RR }
%     & \textbf{(or-nop2)}   %\\
% \end{eqnarray*}
% }
%-

The~$\clear$ function (see Figure~\ref{fig.clear}) concatenates all
active~$\ceu{\Fin}$ bodies of the side being aborted, so that they execute before the
composition rejoins.
Note that there are no transition rules for~$\ceu{\Fin}$ expressions.
This is because once reached, a $\ceu{\Fin}$ expression halts and will only execute 
when it is aborted by a parallel trail and is expanded by the~$\clear$ 
function.
%In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
%the concrete language to a $fin$ in the formal semantics.
%
Note also that there is a syntactic restriction that postulates that~$\ceu{\Fin}$ bodies cannot
contain awaiting expressions ($\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
$\ceu{\Every}$, or $\ceu{\Fin}$),
i.e., the result of a~$\clear$ call is guaranteed to execute entirely within a reaction.

Finally, a~$\ceu{\Break}$ in one of the sides in parallel escapes the closest
enclosing~$\ceu{\Loop}$, properly aborting the other side by applying the~$\clear$
function:
\begin{gather*}
  \hskip-.5em
  \<\ceu{{\Break}\AtAnd{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
  \Rtag{and-brk1}\\[2\jot]
  %%
  \hskip-.5em
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtAnd{\Break}},n,\nil>
    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
  \DisplayProof
  \Rtag{and-brk2}\\[2\jot]
  %%
  \<\ceu{{\Break}\AtOr{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
  \Rtag{or-brk1}\\[2\jot]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtOr{\Break}},n,\nil>
    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
  \DisplayProof
  \Rtag{or-brk2}
\end{gather*}

%-
% { \setlength{\jot}{20pt}
% \begin{eqnarray*}
% & \LL (break~@and~q), n, \epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
%     & \textbf{(and-brk1)}   \\
% %%%
% & \frac
%     { \DS isblocked(n,p) }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@and~break), n, \epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
%     & \textbf{(and-brk2)}   \\
% %%%
% & \LL (break~@or~q),n,\epsilon \RR \NST \LL (clear(q)~;~break),n,\epsilon \RR
%     & \textbf{(or-brk1)}   \\
% %%%
% & \frac
%     { \DS isblocked(n,p) }
% %   -----------------------------------------------------------
%     { \DS \LL (p~@or~break),n,\epsilon \RR \NST \LL (clear(p)~;~break),n,\epsilon \RR }
%     & \textbf{(or-brk2)}   %\\
% \end{eqnarray*}
% }
%-

A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$, $\ceu{\Fin}$,
and~$\ceu{\CanRun}$ expressions in parallel trails.
%
If no trails are blocked in~$\ceu{\CanRun}$ expressions, it means that the program
cannot advance in the current reaction.
%
However, $\ceu{\CanRun}$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \R{pop}).
